---
title: "Running and testing ACMGamp"
author:
  - name: Safa Rafique
    email: safa.sandhu@gmail.com
package: ACMGamp
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Running and testing ACMGamp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = TRUE,
    message = FALSE,
    warning = FALSE
)
```

# Purpose

This vignette explains how to:

1. Launch the ACMGamp Shiny app for interactive testing  
2. Choose sample / benchmark VCFs for each workflow (Groups A–C)  
3. Run automated logic checks and unit tests  

It complements the [Introduction to ACMGamp](ACMGamp.html) vignette.

# Prerequisites

Install CRAN dependencies used by the app:

```{r cran-deps, eval=FALSE}
install.packages(c(
    "shiny", "bslib", "DT", "data.table", "readr", "jsonlite",
    "knitr", "rmarkdown", "testthat", "BiocManager"
))
BiocManager::install(c("BiocStyle", "VariantAnnotation"))
```

Optional (large VCF speed-up on Linux/WSL): install **bcftools** and ensure it is
on `PATH`.

Confirm the package (or source tree) is available:

```{r pkg-check}
pkg_ok <- requireNamespace("ACMGamp", quietly = TRUE)
pkg_ok
```

If you are developing from the Git clone (not yet installed), set the working
directory to the package root (the folder that contains `DESCRIPTION`,
`app.R`, `global.R`, `ui.R`, and `server.R`).

```{r find-root, eval=FALSE}
# Example after cloning:
# setwd("path/to/ACMGamp")   # or cml_variant_interpreter on this machine
```

# Launch the Shiny app

## Preferred Bioconductor API

Package functions **return** a Shiny app; you launch it yourself:

```{r launch-api, eval=FALSE}
library(ACMGamp)

app <- ACMGamp()          # or ACMGampApp()
if (interactive()) {
    shiny::runApp(app, launch.browser = TRUE)
}
```

## Source-tree launch (development)

From the package root, either:

```{r launch-source, eval=FALSE}
# Option 1 — classic Shiny entry
shiny::runApp(".", launch.browser = TRUE)

# Option 2 — after sourcing the package constructor from R/
# (works once ACMGamp is installed or attached)
app <- ACMGamp::ACMGamp()
shiny::runApp(app, launch.browser = TRUE)
```

On Ubuntu/WSL (recommended for large VCFs):

```bash
cd /path/to/ACMGamp
R -e "shiny::runApp('.', host='0.0.0.0', port=3838)"
```

Then open `http://127.0.0.1:3838` in a browser.

## Session settings useful for testing

| Setting | Default | Notes |
|---------|---------|--------|
| Max upload | 1 GiB | Set in `global.R` via `shiny.maxRequestSize` |
| Idle timeout | 1 hour | Override with env `ACMGAMP_IDLE_TIMEOUT_SEC` |
| Auth | Off in local scaffolds when `AUTH_ENABLED <- FALSE` | Re-enable only for secured demos |

```{r env-tips, eval=FALSE}
Sys.setenv(ACMGAMP_IDLE_TIMEOUT_SEC = "3600")
# Optional encryption key for secure uploads (32 characters):
# Sys.setenv(ACMGAMP_ENCRYPTION_KEY = "0123456789abcdef0123456789abcdef")
```

# Test data locations

Paths below assume the **development** repository layout. After full Bioconductor
conversion they will move under `inst/extdata/` and should be loaded with
`system.file()`.

```{r locate-samples}
.root_candidates <- c(
    getwd(),
    file.path(getwd(), "cml_variant_interpreter"),
    system.file(package = "ACMGamp")
)
.root <- .root_candidates[dir.exists(.root_candidates)][1]

sample_vcf <- file.path(.root, "data", "samples", "example_variants.vcf")
clinical_csv <- file.path(.root, "data", "samples", "example_clinical_logs.csv")
pedigree_csv <- file.path(.root, "data", "samples", "example_pedigree.csv")

bench_dir <- normalizePath(
    file.path(.root, "..", "testig", "acmgamp_benchmark"),
    winslash = "/",
    mustWork = FALSE
)
group_b_vcf <- file.path(bench_dir, "acmgamp_group_b_benchmark.vcf")
group_b_key <- file.path(bench_dir, "acmgamp_group_b_benchmark.key.csv")

data.frame(
    resource = c(
        "example_variants.vcf",
        "example_clinical_logs.csv",
        "example_pedigree.csv",
        "group_b_benchmark.vcf",
        "group_b_benchmark.key.csv"
    ),
    path = c(sample_vcf, clinical_csv, pedigree_csv, group_b_vcf, group_b_key),
    exists = file.exists(c(
        sample_vcf, clinical_csv, pedigree_csv, group_b_vcf, group_b_key
    ))
)
```

| File | Use for |
|------|---------|
| `data/samples/example_variants.vcf` | Quick smoke test (few variants) |
| `data/samples/example_clinical_logs.csv` + `example_pedigree.csv` | Group A companion inputs |
| `testig/acmgamp_benchmark/acmgamp_group_b_benchmark.vcf` | Group B accuracy / traps (20 variants) |
| Matching `*.key.csv` / `*.acmg.tsv` | Expected labels for CLI scoring |

# Interactive testing checklist (UI)

## Group B — Automated prediction (start here)

1. Open **Group B — Rapid / Automated**.  
2. Upload `acmgamp_group_b_benchmark.vcf` (preferred) or `example_variants.vcf`.  
3. Confirm readiness checks turn green (CSQ/ANN present).  
4. Preview rows appear; **Run Analysis** is enabled.  
5. Watch the progress bar / status under Run.  
6. Open **Results**: classifications, criteria, confidence.  
7. Download CSV (full report and/or expert worklist).  

**Pass criteria for the 20-variant Group B benchmark** (see
`testig/acmgamp_benchmark/GROUP_B_BENCHMARK.md`):

- Exact accuracy **20/20**  
- All pathogenic anchors detected (TCF4, MECP2, EGR2)  
- No false Pathogenic on benign traps  

## Group C — Gene panel

1. Open **Group C**.  
2. Upload the same VCF as Group B.  
3. Enter a gene list that intersects the call set, for example:

```{r gene-list-example, eval=FALSE}
# Paste into the Group C gene box:
# TCF4, MECP2, EGR2, ATM, TSC1
```

4. Run analysis — only nominated genes should appear in results.  

## Group A — Full clinical

1. Open **Group A**.  
2. Upload VCF + `example_clinical_logs.csv` + `example_pedigree.csv`.  
3. Optionally enable curator toggles (PS2, PS3, PS4, PM6, PP1, PP4).  
4. Run and confirm criteria appear in evidence export.  

# Automated testing (command line)

Run these from the package root (`DESCRIPTION` present).

## Unit tests (testthat)

```{r unit-tests, eval=FALSE}
# After install:
# BiocManager::install("safarafique/ACMGamp")  # or local install
devtools::test()           # development
# or:
testthat::test_package("ACMGamp")
```

Minimal checks included in `tests/testthat/test-ACMGamp.R` verify that
`ACMGamp()` / `ACMGampApp()` exist and return a `shiny.appobj` when Shiny entry
files are present.

## Engine / pipeline scripts

```{r cli-scripts, eval=FALSE}
# Rules combining + criterion logic audit
Rscript scripts/audit_acmg_rules.R

# Group B/C coverage and gene-filter parity
Rscript scripts/verify_group_b_c.R

# Group A 28-criterion catalog check
Rscript scripts/verify_group_a_28.R

# Score the Group B benchmark VCF vs expected labels
Rscript scripts/run_group_b_benchmark.R

# Smaller accuracy fixture
Rscript scripts/run_accuracy_benchmark.R

# Formal validation report (expert set)
Rscript scripts/generate_validation_report.R
```

Quick existence check for key scripts (does not execute them in the vignette
build):

```{r script-list}
scripts <- c(
    "audit_acmg_rules.R",
    "verify_group_b_c.R",
    "verify_group_a_28.R",
    "run_group_b_benchmark.R",
    "run_accuracy_benchmark.R",
    "generate_validation_report.R",
    "test_upload_flow.R"
)
script_paths <- file.path(.root, "scripts", scripts)
data.frame(script = scripts, exists = file.exists(script_paths))
```

## Optional upload-flow smoke script

```{r upload-flow, eval=FALSE}
Rscript scripts/test_upload_flow.R
```

## Optional Bioconductor / style checks (before submission)

```{r bioc-check, eval=FALSE}
# From the parent of the package directory:
# R CMD build ACMGamp
# R CMD check ACMGamp_0.99.0.tar.gz
# BiocCheck::BiocCheck("ACMGamp_0.99.0.tar.gz", `new-package` = TRUE)
# BiocCheck::BiocCheckGitClone("ACMGamp")
```

# Expected app object (non-interactive sanity check)

This chunk verifies that the constructor returns a Shiny app when package
sources are on the search path or the working directory is the package root:

```{r app-object}
has_entry <- file.exists(file.path(.root, "global.R")) &&
    file.exists(file.path(.root, "ui.R")) &&
    file.exists(file.path(.root, "server.R"))

if (requireNamespace("ACMGamp", quietly = TRUE) && has_entry) {
    old <- getwd()
    on.exit(setwd(old), add = TRUE)
    setwd(.root)
    app <- ACMGamp::ACMGamp()
    inherits(app, "shiny.appobj")
} else if (has_entry) {
    "# Install/attach ACMGamp or run shiny::runApp('.') from package root"
} else {
    "# Shiny entry files not found — setwd() to package root first"
}
```

# Troubleshooting while testing

| Symptom | What to check |
|---------|----------------|
| Upload stuck / empty validation | Restart app; confirm VCF has `CSQ` or `ANN`; wait for readiness UI |
| Run does nothing | Use Group B static Run button; look for status / progress under button; check R console for `[Group B] Run clicked` |
| Auth modal blocks use | Set `AUTH_ENABLED <- FALSE` in `R/auth_storage.R` for local testing |
| AES-GCM encrypt error | Use 12-byte IV build; set `ACMGAMP_ENCRYPTION_KEY` (32 chars) if encryption on |
| High VUS rate on WGS | Expected without gene filter — try Group C |
| Package not found | `setwd()` to package root or `devtools::load_all()` / install |

# Recommended smoke sequence (15 minutes)

1. `shiny::runApp(".")`  
2. Group B + `acmgamp_group_b_benchmark.vcf` → Run → download CSV  
3. Group C with `TCF4,MECP2,EGR2` → confirm filtered genes  
4. Group A with example CSV companions → Run  
5. `Rscript scripts/audit_acmg_rules.R`  
6. `Rscript scripts/run_group_b_benchmark.R`  

# Session info

```{r sessionInfo}
sessionInfo()
```
