---
title: "Introduction to ClinicalVariantR"
author:
  - name: Safa Rafique
    email: safa.sandhu@gmail.com
    orcid: 0000-0003-2646-8106
  - name: Naeem Mahmood
    email: naeem.sbb@pu.edu.pk
  - name: Muhammad Farooq Sabar
    email: farooq.sbb@pu.edu.pk
package: ClinicalVariantR
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to ClinicalVariantR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = TRUE,
    message = FALSE,
    warning = FALSE
)
```

# Overview

**ClinicalVariantR** is an R / Bioconductor software package that classifies germline
sequence variants under the ACMG/AMP 2015 framework from pre-annotated VCF
files (VEP CSQ, SnpEff ANN, or ANNOVAR-style INFO fields).

Source code:
[https://github.com/safarafique/ClinicalVariantR](https://github.com/safarafique/ClinicalVariantR)

The package exposes an interactive Shiny application with three workflows:

| Workflow | Inputs | Purpose |
|----------|--------|---------|
| Group A (full clinical) | VCF + clinical log + pedigree | Automated criteria plus curator toggles |
| Group B (rapid) | VCF only | High-throughput automated classification |
| Group C (gene panel) | VCF + gene list | Panel-restricted automated classification |

# Install and open the Shiny app

The package ships a complete Shiny app under `inst/shinyapp/`
(`global.R`, `ui.R`, `server.R`, plus `config/`, `data/`, and `www/`).
`ClinicalVariantR()` locates that directory after a successful install.

## Recommended: install from local clone (Windows)

On Windows, **restart R before installing** if `ClinicalVariantR` is already
loaded. Otherwise you may see
`package 'ClinicalVariantR' is in use and will not be installed` and keep the
old launcher.

```r
# 0. Session -> Restart R  (required if the package is already loaded)

# 1. Install from the local package folder
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}
remotes::install_local(
    "E:/ACGM/ClinicalVariantR",
    force = TRUE,
    upgrade = "never"
)

# 2. Load and open the app
library(ClinicalVariantR)
app <- ClinicalVariantR()   # or ClinicalVariantRApp()
if (interactive()) {
    shiny::runApp(app, launch.browser = TRUE)
}
```

## Install from GitHub

Until the package is accepted on Bioconductor, you can also install from GitHub
(use this after the fixed launcher has been pushed):

```r
# Session -> Restart R first if ClinicalVariantR is already loaded
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}
remotes::install_github("safarafique/ClinicalVariantR")

library(ClinicalVariantR)
app <- ClinicalVariantR()
if (interactive()) {
    shiny::runApp(app, launch.browser = TRUE)
}
```

## Fast path: run the source tree (no reinstall)

If you only need to open the app from a local checkout:

```r
shiny::runApp("E:/ACGM/ClinicalVariantR", launch.browser = TRUE)
```

Following Bioconductor Shiny guidelines, package functions **return** a Shiny
app object; you call `shiny::runApp()` yourself. Do not expect
`ClinicalVariantR()` alone to open a browser window.

## Optional Bioconductor install (after acceptance)

```r
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("ClinicalVariantR")
library(ClinicalVariantR)
app <- ClinicalVariantR()
if (interactive()) {
    shiny::runApp(app, launch.browser = TRUE)
}
```

## Sanity check (evaluated in this vignette)

This chunk confirms the constructor is available when the package is installed.
It does **not** launch the browser during vignette building.

```{r launch}
if (requireNamespace("ClinicalVariantR", quietly = TRUE)) {
    stopifnot(is.function(ClinicalVariantR::ClinicalVariantR))
    stopifnot(is.function(ClinicalVariantR::ClinicalVariantRApp))
    # Interactive use (not run during vignette builds):
    # shiny::runApp(ClinicalVariantR::ClinicalVariantR(), launch.browser = TRUE)
}
```

For sample VCFs, Group A/B/C UI checks, and command-line tests, see the
companion vignette **Running and testing ClinicalVariantR**.

# Input requirements

- **VCF** (`.vcf` or `.vcf.gz`) with VEP `CSQ`, SnpEff `ANN`, or ANNOVAR-style
  annotation for consequence, gene symbol, transcript, population AF, ClinVar,
  and in silico scores when available.
- Optional **clinical log** and **pedigree** CSV files for Group A.
- Optional **gene symbol list** for Group C.

Default reference tables under `inst/extdata/reference/` are placeholders;
population and prediction evidence primarily use annotation embedded in the
VCF unless curated references are installed (see package `README`).

# Scope and limitations

ClinicalVariantR targets **germline SNV and small-indel** interpretation. It does not
classify somatic variants, structural variants, CNVs, or mitochondrial genomes.
Several ACMG criteria require external data (functional assays, phasing,
matched pathogenic variants) and are registered as manual or not automated.

See the package `NEWS` file and GitHub README for version-specific criterion
coverage and validation notes.

# Session info

```{r sessionInfo}
sessionInfo()
```
