---
title: "Running and testing ClinicalVariantR"
author:
  - name: Safa Rafique
    email: safa.sandhu@gmail.com
    orcid: 0000-0003-2646-8106
  - name: Naeem Mahmood
    email: naeem.sbb@pu.edu.pk
  - name: Muhammad Farooq Sabar
    email: farooq.sbb@pu.edu.pk
package: ClinicalVariantR
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Running and testing ClinicalVariantR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = TRUE,
    message = FALSE,
    warning = FALSE
)
```

# Purpose

This vignette explains how to:

1. Install ClinicalVariantR and open the Shiny app
2. Choose sample / benchmark VCFs for each workflow (Groups A-C)
3. Run automated logic checks and unit tests

It complements the [Introduction to ClinicalVariantR](ClinicalVariantR.html) vignette.

Repository:
[https://github.com/safarafique/ClinicalVariantR](https://github.com/safarafique/ClinicalVariantR)

# Install and open the Shiny app

## Working Windows workflow (verified)

If `ClinicalVariantR` is already loaded, RStudio will refuse to overwrite it
(`package 'ClinicalVariantR' is in use and will not be installed`). Always
**Session → Restart R** first.

```r
# 0. Session -> Restart R

# 1. Install from the local package folder
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}
remotes::install_local(
    "E:/ACGM/ClinicalVariantR",
    force = TRUE,
    upgrade = "never"
)

# 2. Open the app (Bioconductor-style API)
library(ClinicalVariantR)
app <- ClinicalVariantR()   # or ClinicalVariantRApp()
if (interactive()) {
    shiny::runApp(app, launch.browser = TRUE)
}
```

After a successful install, `ClinicalVariantR()` uses the bundled app in
`inst/shinyapp/` (`global.R`, `ui.R`, `server.R`, `config/`, `data/`, `www/`).

## Install from GitHub

```r
# Session -> Restart R first if the package is already loaded
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}
remotes::install_github("safarafique/ClinicalVariantR")

library(ClinicalVariantR)
app <- ClinicalVariantR()
if (interactive()) {
    shiny::runApp(app, launch.browser = TRUE)
}
```

## Fast path: source-tree launch (no reinstall)

```r
shiny::runApp("E:/ACGM/ClinicalVariantR", launch.browser = TRUE)
```

Optional CRAN / Bioconductor dependencies used by the app and docs:

```r
install.packages(c(
    "shiny", "bslib", "DT", "data.table", "readr", "jsonlite",
    "knitr", "rmarkdown", "testthat", "BiocManager", "remotes"
))
BiocManager::install(c("BiocStyle", "VariantAnnotation"))
```

Optional (large VCF speed-up on Linux/WSL): install **bcftools** and ensure it is
on `PATH`.

Confirm the package is available after install:

```{r pkg-check}
pkg_ok <- requireNamespace("ClinicalVariantR", quietly = TRUE)
pkg_ok
```

## Preferred Bioconductor API (sanity check)

Package functions **return** a Shiny app; you launch it yourself. This chunk
only checks that the constructor exists during vignette builds:

```{r launch-api}
if (requireNamespace("ClinicalVariantR", quietly = TRUE)) {
    stopifnot(is.function(ClinicalVariantR::ClinicalVariantR))
    stopifnot(is.function(ClinicalVariantR::ClinicalVariantRApp))
}
# Interactive launch (copy into an R console):
# library(ClinicalVariantR)
# shiny::runApp(ClinicalVariantR(), launch.browser = TRUE)
```

## Source-tree / clone launch (development)

If you are already inside the package root (folder with `DESCRIPTION`,
`app.R`, `global.R`, `ui.R`, and `server.R`):

```{r find-root}
normalizePath(getwd(), winslash = "/", mustWork = FALSE)
```

```r
# Classic Shiny entry from the package root
shiny::runApp(".", launch.browser = TRUE)

# Or absolute path (works from any working directory)
shiny::runApp("E:/ACGM/ClinicalVariantR", launch.browser = TRUE)
```

On Ubuntu/WSL (recommended for large VCFs):

```bash
cd /path/to/ClinicalVariantR
R -e "shiny::runApp('.', host='0.0.0.0', port=3838)"
```

Then open `http://127.0.0.1:3838` in a browser.

## Session settings useful for testing

| Setting | Default | Notes |
|---------|---------|--------|
| Max upload | 1 GiB | Set in `global.R` via `shiny.maxRequestSize` |
| Idle timeout | 1 hour | Override with env `CLINICALVARIANTR_IDLE_TIMEOUT_SEC` |
| Auth | Off in local scaffolds when `AUTH_ENABLED <- FALSE` | Re-enable only for secured demos |

```{r env-tips}
Sys.setenv(CLINICALVARIANTR_IDLE_TIMEOUT_SEC = "3600")
Sys.getenv("CLINICALVARIANTR_IDLE_TIMEOUT_SEC")
```

# Test data locations

Paths below assume the **development** repository layout. After full Bioconductor
conversion they will move under `inst/extdata/` and should be loaded with
`system.file()`.

```{r locate-samples}
.root_candidates <- c(
    getwd(),
    file.path(getwd(), "ClinicalVariantR"),
    system.file(package = "ClinicalVariantR")
)
.root <- .root_candidates[dir.exists(.root_candidates)][1]

sample_vcf <- file.path(.root, "data", "samples", "example_variants.vcf")
clinical_csv <- file.path(.root, "data", "samples", "example_clinical_logs.csv")
pedigree_csv <- file.path(.root, "data", "samples", "example_pedigree.csv")

bench_dir <- normalizePath(
    file.path(.root, "..", "testig", "clinicalvariantr_benchmark"),
    winslash = "/",
    mustWork = FALSE
)
group_b_vcf <- file.path(bench_dir, "clinicalvariantr_group_b_benchmark.vcf")
group_b_key <- file.path(bench_dir, "clinicalvariantr_group_b_benchmark.key.csv")

data.frame(
    resource = c(
        "example_variants.vcf",
        "example_clinical_logs.csv",
        "example_pedigree.csv",
        "group_b_benchmark.vcf",
        "group_b_benchmark.key.csv"
    ),
    path = c(sample_vcf, clinical_csv, pedigree_csv, group_b_vcf, group_b_key),
    exists = file.exists(c(
        sample_vcf, clinical_csv, pedigree_csv, group_b_vcf, group_b_key
    ))
)
```

| File | Use for |
|------|---------|
| `data/samples/example_variants.vcf` | Quick smoke test (few variants) |
| `data/samples/example_clinical_logs.csv` + `example_pedigree.csv` | Group A companion inputs |
| `testig/clinicalvariantr_benchmark/clinicalvariantr_group_b_benchmark.vcf` | Group B accuracy / traps (20 variants) |
| Matching `*.key.csv` / `*.acmg.tsv` | Expected labels for CLI scoring |

# Interactive testing checklist (UI)

## Group B - Automated prediction (start here)

1. Open **Group B - Rapid / Automated**.
2. Upload `clinicalvariantr_group_b_benchmark.vcf` (preferred) or `example_variants.vcf`.
3. Confirm readiness checks turn green (CSQ/ANN present).
4. Preview rows appear; **Run Analysis** is enabled.
5. Watch the progress bar / status under Run.
6. Open **Results**: classifications, criteria, confidence.
7. Download CSV (full report and/or expert worklist).

**Pass criteria for the 20-variant Group B benchmark** (see
`testig/clinicalvariantr_benchmark/GROUP_B_BENCHMARK.md`):

- Exact accuracy **20/20**
- All pathogenic anchors detected (TCF4, MECP2, EGR2)
- No false Pathogenic on benign traps

## Group C - Gene panel

1. Open **Group C**.
2. Upload the same VCF as Group B.
3. Enter a gene list that intersects the call set, for example:

```r
# Paste into the Group C gene box:
# TCF4, MECP2, EGR2, ATM, TSC1
```

4. Run analysis - only nominated genes should appear in results.

## Group A - Full clinical

1. Open **Group A**.
2. Upload VCF + `example_clinical_logs.csv` + `example_pedigree.csv`.
3. Optionally enable curator toggles (PS2, PS3, PS4, PM6, PP1, PP4).
4. Run and confirm criteria appear in evidence export.

# Automated testing (command line)

Run these from the package root (`DESCRIPTION` present).

## Unit tests (testthat)

```r
# After install:
# BiocManager::install("safarafique/ClinicalVariantR")  # or local install
devtools::test()           # development
# or:
testthat::test_package("ClinicalVariantR")
```

Minimal checks included in `tests/testthat/test-ClinicalVariantR.R` verify that
`ClinicalVariantR()` / `ClinicalVariantRApp()` exist and return a `shiny.appobj` when Shiny entry
files are present.

## Engine / pipeline scripts

```r
# Rules combining + criterion logic audit
Rscript scripts/audit_acmg_rules.R

# Group B/C coverage and gene-filter parity
Rscript scripts/verify_group_b_c.R

# Group A 28-criterion catalog check
Rscript scripts/verify_group_a_28.R

# Score the Group B benchmark VCF vs expected labels
Rscript scripts/run_group_b_benchmark.R

# Smaller accuracy fixture
Rscript scripts/run_accuracy_benchmark.R

# Formal validation report (expert set)
Rscript scripts/generate_validation_report.R
```

Quick existence check for key scripts (does not execute them in the vignette
build):

```{r script-list}
scripts <- c(
    "audit_acmg_rules.R",
    "verify_group_b_c.R",
    "verify_group_a_28.R",
    "run_group_b_benchmark.R",
    "run_accuracy_benchmark.R",
    "generate_validation_report.R",
    "test_upload_flow.R"
)
script_paths <- file.path(.root, "scripts", scripts)
data.frame(script = scripts, exists = file.exists(script_paths))
```

## Optional upload-flow smoke script

```r
Rscript scripts/test_upload_flow.R
```

## Optional Bioconductor / style checks (before submission)

```r
# From the parent of the package directory:
# R CMD build ClinicalVariantR
# R CMD check ClinicalVariantR_0.99.0.tar.gz
# BiocCheck::BiocCheck("ClinicalVariantR_0.99.0.tar.gz", `new-package` = TRUE)
# BiocCheck::BiocCheckGitClone("ClinicalVariantR")
```

# Expected app object (non-interactive sanity check)

This chunk verifies that the constructor returns a Shiny app when package
sources are on the search path or the working directory is the package root:

```{r app-object}
has_entry <- file.exists(file.path(.root, "global.R")) &&
    file.exists(file.path(.root, "ui.R")) &&
    file.exists(file.path(.root, "server.R"))

if (requireNamespace("ClinicalVariantR", quietly = TRUE) && has_entry) {
    old <- getwd()
    on.exit(setwd(old), add = TRUE)
    setwd(.root)
    app <- ClinicalVariantR::ClinicalVariantR()
    inherits(app, "shiny.appobj")
} else if (has_entry) {
    "# Install/attach ClinicalVariantR or run shiny::runApp('.') from package root"
} else {
    "# Shiny entry files not found - setwd() to package root first"
}
```

# Troubleshooting while testing

| Symptom | What to check |
|---------|----------------|
| Upload stuck / empty validation | Restart app; confirm VCF has `CSQ` or `ANN`; wait for readiness UI |
| Run does nothing | Use Group B static Run button; look for status / progress under button; check R console for `[Group B] Run clicked` |
| Auth modal blocks use | Set `AUTH_ENABLED <- FALSE` in `R/auth_storage.R` for local testing |
| AES-GCM encrypt error | Use 12-byte IV build; set `CLINICALVARIANTR_ENCRYPTION_KEY` (32 chars) if encryption on |
| High VUS rate on WGS | Expected without gene filter - try Group C |
| Package not found | `setwd()` to package root or `devtools::load_all()` / install |

# Recommended smoke sequence (15 minutes)

1. Restart R, then
   `remotes::install_local("E:/ACGM/ClinicalVariantR", force = TRUE, upgrade = "never")`
   and `shiny::runApp(ClinicalVariantR(), launch.browser = TRUE)`
   (or `shiny::runApp("E:/ACGM/ClinicalVariantR", launch.browser = TRUE)`)
2. Group B + `clinicalvariantr_group_b_benchmark.vcf` -> Run -> download CSV
3. Group C with `TCF4,MECP2,EGR2` -> confirm filtered genes
4. Group A with example CSV companions -> Run
5. `Rscript scripts/audit_acmg_rules.R`
6. `Rscript scripts/run_group_b_benchmark.R`

# Session info

```{r sessionInfo}
sessionInfo()
```
