---
title: "Introduction to ClinicalVariantR"
author:
  - name: Safa Rafique
    affiliation: University of the Punjab
  - name: Naeem Mahmood
    affiliation: University of the Punjab
  - name: Muhammad Farooq Sabar
    affiliation: University of the Punjab
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to ClinicalVariantR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

# Introduction

`ClinicalVariantR` is a **Shiny application** for evidence-based germline
variant classification under the ACMG/AMP 2015 framework. This vignette
walks through installing the package, launching the app, and using the
three prediction workflows — **Group A** (clinical + curator), **Group B**
(rapid automated), and **Group C** (gene-panel automated) — from annotated
VCF files (VEP `CSQ`, SnpEff `ANN`, or ANNOVAR-style INFO fields).

## Vignette outline

| Section | Purpose |
|------------------------------------|------------------------------------|
| [How the Shiny app is organised](#how-the-shiny-app-is-organised) | Workflow overview figure |
| [Installation](#installation) | Bioconductor (submitted) vs GitHub (not yet) vs local |
| [Run latest code for analysis](#run-latest-code) | Avoid stale installs; reinstall / source launch |
| [Launch the Shiny application](#launch-the-shiny-application) | Start the app + [app steps](#app-steps) |
| [Evaluated package checks](#evaluated-package-checks) | Constructors, samples, parser demos |
| [Step-by-step Shiny walkthrough](#step-by-step-shiny-walkthrough) | Home and Group A / B / C actions |
| [VCF annotation requirements](#vcf-annotation-requirements) | CSQ / ANN / ANNOVAR expectations |
| [Sample inputs](#sample-inputs) | Bundled example files |
| [Troubleshooting](#troubleshooting) | Common install and runtime issues |

The package builds on `r BiocStyle::CRANpkg("shiny")` and
`r BiocStyle::CRANpkg("bslib")` for the interface,
`r BiocStyle::CRANpkg("data.table")` for streaming tabular work, and
`r BiocStyle::Biocpkg("VariantAnnotation")` for VCF handling.

## How the Shiny app is organised {#how-the-shiny-app-is-organised}

The top navbar lists **Home**, **Group A**, **Group B**, **Group C**, plus
optional **Explorer** and **Audit** panels. Start on **Home**, pick a
workflow card, upload inputs, wait for readiness checks, then click
**Run Analysis**.

| Phase | Tabs | What you do |
|------------------------|------------------------|------------------------|
| **Orientation** | Home | Choose Group A, B, or C |
| **Rapid prediction** | Group B | Upload VCF → validate → run → download |
| **Panel prediction** | Group C | Upload VCF + gene list → run filtered results |
| **Clinical curation** | Group A | Upload VCF + clinical + pedigree → curator toggles → run |
| **Review** | Explorer / Audit | Browse evidence and session metadata after a run |

```{r pipeline-overview, echo=FALSE, fig.cap="ClinicalVariantR Shiny workflows. Each box corresponds to a navbar tab or analysis phase."}
workflow <- data.frame(
  step  = 1:6,
  phase = c(
    "Orientation",
    "Rapid prediction",
    "Panel prediction",
    "Clinical curation",
    "Review",
    "Review"
  ),
  label = c(
    "1 Home",
    "2 Group B",
    "3 Group C",
    "4 Group A",
    "5 Explorer",
    "6 Audit"
  ),
  stringsAsFactors = FALSE
)
phase_cols <- c(
  "Orientation"        = "#4E79A7",
  "Rapid prediction"   = "#59A14F",
  "Panel prediction"   = "#F28E2B",
  "Clinical curation"  = "#E15759",
  "Review"             = "#76B7B2"
)
workflow$y <- rev(workflow$step)

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(workflow, ggplot2::aes(x = 1, y = y, fill = phase)) +
    ggplot2::geom_tile(
      ggplot2::aes(width = 0.92, height = 0.88),
      colour = "white", linewidth = 0.4
    ) +
    ggplot2::geom_text(
      ggplot2::aes(label = label),
      colour = "white", size = 3.4, fontface = "bold"
    ) +
    ggplot2::scale_fill_manual(values = phase_cols, name = "Phase") +
    ggplot2::scale_x_continuous(limits = c(0.5, 1.5), expand = c(0, 0)) +
    ggplot2::scale_y_continuous(expand = c(0.02, 0.02)) +
    ggplot2::labs(x = NULL, y = NULL, title = "Shiny app workflow") +
    ggplot2::theme_void() +
    ggplot2::theme(
      legend.position = "bottom",
      plot.title = ggplot2::element_text(hjust = 0.5, face = "bold")
    )
} else {
  knitr::kable(
    workflow[, c("label", "phase")],
    caption = "ClinicalVariantR Shiny workflow (install ggplot2 for the coloured overview figure)."
  )
}
```

# Installation {#installation}

Follow the [Bioconductor installation
instructions](https://contributions.bioconductor.org/docs.html#installation)
and the [official install guide](https://bioconductor.org/install/) so your
R version matches a supported Bioconductor release (e.g. Bioconductor 3.22
for R 4.6). ClinicalVariantR requires **R ≥ 4.6.0**.

Always install with **`dependencies = TRUE`** so Imports (shiny, bslib, DT,
data.table, readr, jsonlite, VariantAnnotation, …) are resolved once. The
app does **not** install packages at launch.

Pick the path that matches the package status:

| Status | Use this install |
|--------|------------------|
| **Accepted / available on Bioconductor** | Section [A — Bioconductor](#install-bioconductor) |
| **Not yet on Bioconductor** (current GitHub release) | Section [B — GitHub](#install-github) |
| **Local development clone** | Section [C — Local clone](#install-local) |

If Windows reports that the package is in use, use **Session → Restart R**
before reinstalling or updating.

## A — Bioconductor (when the package is submitted and available) {#install-bioconductor}

Use this after ClinicalVariantR is listed on Bioconductor (or when
reinstalling / updating from the Bioconductor build):

```r
# Session -> Restart R if ClinicalVariantR is already loaded

if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("ClinicalVariantR", dependencies = TRUE)

library(ClinicalVariantR)
packageVersion("ClinicalVariantR")  # expect 0.99.3
```

To **reinstall / recall** the Bioconductor build after an update:

```r
# Session -> Restart R first
BiocManager::install("ClinicalVariantR", force = TRUE, dependencies = TRUE)
library(ClinicalVariantR)
```

## B — GitHub (when not yet on Bioconductor) {#install-github}

Until the package is on Bioconductor, install (or reinstall) from
[GitHub](https://github.com/safarafique/ClinicalVariantR):

```r
# Session -> Restart R if ClinicalVariantR is already loaded
options(timeout = 3600)

if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
options(repos = BiocManager::repositories())

if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}

remotes::install_github(
    "safarafique/ClinicalVariantR",
    ref = "main",
    dependencies = TRUE,
    force = TRUE,
    INSTALL_opts = "--no-staged-install"
)

library(ClinicalVariantR)
packageVersion("ClinicalVariantR")  # expect 0.99.3
```

## C — Local clone (developers) {#install-local}

```r
# Session -> Restart R first
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}
remotes::install_local(
    "E:/ACGM/ClinicalVariantR",
    force = TRUE,
    upgrade = "never",
    dependencies = TRUE
)
library(ClinicalVariantR)
```

Or launch the source tree without reinstalling (see
[Run latest code for analysis](#run-latest-code)).

# Run latest code for analysis {#run-latest-code}

If you installed some time ago, reinstall with section A or B above so
`inst/shinyapp/` is current, or launch from a local clone:

```r
# Local clone — no reinstall
shiny::runApp("E:/ACGM/ClinicalVariantR", launch.browser = TRUE)
```

```r
library(ClinicalVariantR)
app <- ClinicalVariantR()
shiny::runApp(app, launch.browser = TRUE)
```

# Launch the Shiny application {#launch-the-shiny-application}

Bioconductor Shiny apps return an app object from the package; users
start the server with `shiny::runApp()` ([Bioconductor Shiny
guide](https://contributions.bioconductor.org/shiny.html#documentation)).

After install (Bioconductor **or** GitHub):

```r
library(ClinicalVariantR)
app <- ClinicalVariantR()
shiny::runApp(app, launch.browser = TRUE)
```

`ClinicalVariantR()` **returns** a Shiny app object; it does not open a
browser by itself. `shiny::runApp(app, launch.browser = TRUE)` starts the
server and opens your default browser. `ClinicalVariantRApp()` is an alias
of `ClinicalVariantR()`.

## App steps at a glance {#app-steps}

After the browser opens:

1. **Home** — read the three workflow cards and pick a mode.
2. **Group B (recommended first run)** — upload annotated VCF → wait for
   readiness → **Run Analysis** → download CSV.
3. **Group C** — upload VCF → paste gene symbols → **Run Analysis** →
   confirm panel-restricted results.
4. **Group A** — upload VCF + clinical CSV + pedigree CSV → optional
   curator toggles → **Run Analysis** → review 28-criteria evidence →
   export.
5. **Explorer / Audit** — after any successful run, browse evidence and
   session metadata.

| Step | Tab | Inputs | Action |
|------|-----|--------|--------|
| 1 | Home | — | Choose Group A, B, or C |
| 2 | Group B | Annotated VCF | Validate → Run → download |
| 3 | Group C | VCF + gene list | Validate → Run → filtered table |
| 4 | Group A | VCF + clinical + pedigree | Validate → Run → curator / export |
| 5 | Explorer / Audit | Prior run results | Review evidence and logs |

Detailed click-by-click instructions are in
[Step-by-step Shiny walkthrough](#step-by-step-shiny-walkthrough).

## Evaluated package checks {#evaluated-package-checks}

The chunks below run during vignette builds. They do **not** open a browser.

```{r load-pkg}
library(ClinicalVariantR)
packageVersion("ClinicalVariantR")
```

```{r launch-check}
stopifnot(is.function(ClinicalVariantR))
stopifnot(is.function(ClinicalVariantRApp))
app <- ClinicalVariantR()
stopifnot(inherits(app, "shiny.appobj"))
class(app)
```

```{r criteria-constants}
length(ClinicalVariantR:::AUTOMATED_ACMG_CRITERIA)
ClinicalVariantR:::AUTOMATED_ACMG_CRITERIA
ClinicalVariantR:::ACMG_CLASSIFICATIONS
```

```{r gene-filter-demo}
ClinicalVariantR:::parse_gene_filter("TCF4, MECP2; EGR2\nBCR-ABL")
```

```{r evidence-combine-demo}
ClinicalVariantR:::combine_acmg_evidence(list(
  PVS = 1L, PS = 0L, PM = 1L, PP = 1L,
  BA = FALSE, BS = 0L, BP = 0L
))
ClinicalVariantR:::normalize_acmg_class("Likely_pathogenic")
```

```{r vcf-requirements}
ClinicalVariantR:::vcf_app_requirements("rapid")
ClinicalVariantR:::vcf_app_requirements("full")
```

# Step-by-step Shiny walkthrough {#step-by-step-shiny-walkthrough}

The sections below mirror the navbar tabs. For each workflow, read the
**goal**, apply the settings under **What to do**, click **Run Analysis**,
and confirm the **success check** before exporting results.

## Home — choose a workflow

**Navbar:** `Home`

**Goal:** Orient yourself and open the right prediction mode.

**What to do:**

1. Confirm the app title **ClinicalVariantR — Variant Prediction Platform**.
2. Read the three landing cards:
   - **Group A — Clinical Prediction** (VCF + clinical + pedigree; 28 criteria)
   - **Group B — Automated Prediction** (VCF only; 18 automated criteria)
   - **Group C — Gene Panel Prediction** (VCF + gene list; same engine as B)
3. Click a card (or the matching navbar tab) to continue.

**Success check:** The selected Group tab opens with upload controls on the
left and validation / results panels on the right.

## Group B — Automated Prediction (start here)

**Navbar:** `Group B — Automated Prediction`

**Goal:** Fastest smoke test — classify an annotated VCF with automated
ACMG/AMP rules only.

**What to do:**

1. Under **Required Inputs**, upload a VCF (`.vcf` / `.vcf.gz`), for example
   `data/samples/example_variants.vcf` or a Group B benchmark VCF.
2. Optionally set disease profile, **Analyze entire VCF**, PASS-only filter,
   minimum QUAL, chunk size, and bcftools (Ubuntu/WSL).
3. Wait for **VCF Requirement Check** to report ready (CSQ / ANN / required
   fields present).
4. Review the **VCF Preview** table.
5. Click **Run Analysis**.
6. Open **Classification Results**: summary counts, variant table, criteria /
   confidence.
7. Download CSV (full report and/or expert worklist) when offered.

**Success check:** Readiness is green; results populate with Pathogenic /
Likely Pathogenic / VUS / Likely Benign / Benign categories; CSV download
works.

## Group C — Gene Panel Prediction

**Navbar:** `Group C — Gene Panel Prediction`

**Goal:** Same automated engine as Group B, restricted to nominated genes.

**What to do:**

1. Upload the patient VCF.
2. Paste HGNC gene symbols into **Gene symbol(s)**, for example:

```r
# Paste into the Group C gene box:
# TCF4, MECP2, EGR2, ATM, TSC1
```

3. Confirm readiness, then click **Run Analysis**.
4. Verify that results are limited to the nominated gene set.

**Success check:** Only nominated genes appear in the filtered result table.

## Group A — Clinical Prediction

**Navbar:** `Group A — Clinical Prediction`

**Goal:** Full clinical mode with clinical logs, pedigree, and per-variant
curator evidence (28-criteria reporting).

**What to do:**

1. Upload **VCF**, **Clinical Logs (CSV)**, and **Pedigree Data (CSV)**.
   Bundled examples:
   - `data/samples/example_variants.vcf`
   - `data/samples/example_clinical_logs.csv`
   - `data/samples/example_pedigree.csv`
2. Wait until Group A readiness reports all three inputs validated.
3. Optionally enable per-variant curator evidence after a first pass (e.g.
   PS2, PS3, PS4, PM6, PP1, PP4) in the evidence accordion.
4. Click **Run Analysis**.
5. Select a variant row to inspect 28-criteria detail; apply curator toggles
   and reclassify when appropriate.
6. Export evidence / CSV downloads.

**Success check:** All three inputs validate; classification and evidence
panels populate; curator overrides update the selected variant.

## Explorer and Audit (after a run)

**Navbar:** `Explorer` / `Audit`

**Goal:** Browse structured evidence and session metadata for completed
analyses.

**What to do:**

1. Complete at least one Group A, B, or C run.
2. Open **Explorer** for report-style summaries and variant evidence.
3. Open **Audit** for reproducibility / session logging (unless skip-audit
   was enabled).

**Success check:** Explorer is not empty after a successful run.

# VCF annotation requirements {#vcf-annotation-requirements}

ClinicalVariantR expects a **pre-annotated** VCF. At least one of the
following should be present:

| Annotation style | Typical INFO / field | Notes |
|------------------|----------------------|-------|
| VEP | `CSQ` | Preferred for transcript consequences |
| SnpEff | `ANN` | Accepted alternative |
| ANNOVAR-style | INFO key/value fields | Supported when required columns resolve |

Optional but helpful fields (when available) improve automated criteria:
population frequencies, computational predictions, ClinVar assertions, and
gene / consequence annotations used by PVS1 and related rules.

Group A additionally requires:

- **Clinical logs CSV** — phenotype / clinical context for curator-assisted
  criteria
- **Pedigree CSV** — family structure for segregation-related evidence

# Sample inputs {#sample-inputs}

Installed packages ship samples under `inst/extdata/samples/` and
`inst/shinyapp/data/samples/`. Development clones may still have a package-root
`data/samples/` tree (ignored by the package build).

```{r locate-samples}
extdata_samples <- system.file("extdata", "samples", package = "ClinicalVariantR")
sample_files <- c(
  "example_variants.vcf",
  "example_clinical_logs.csv",
  "example_pedigree.csv"
)
sample_paths <- file.path(extdata_samples, sample_files)
data.frame(
  resource = sample_files,
  path = sample_paths,
  exists = file.exists(sample_paths)
)
```

```{r preview-example-vcf}
vcf_path <- system.file(
  "extdata", "samples", "example_variants.vcf",
  package = "ClinicalVariantR"
)
stopifnot(nzchar(vcf_path), file.exists(vcf_path))
writeLines(readLines(vcf_path, n = 8L))
preview <- ClinicalVariantR:::preview_vcf(vcf_path, max_rows = 5L)
preview$total_variants
preview$preview
```

```{r parse-vcf-line}
hdr <- c("CHROM", "POS", "ID", "REF", "ALT", "QUAL", "FILTER", "INFO")
line <- "chr9\t133738352\t.\tG\tA\t.\tPASS\tGENE=ABL1;CSQ=stop_gained;AF=0.00045;REVEL=0.91"
ClinicalVariantR:::parse_vcf_line(line, hdr)
```

| File | Use for |
|------|---------|
| `example_variants.vcf` | Quick smoke test (few variants) |
| `example_clinical_logs.csv` | Group A clinical companion |
| `example_pedigree.csv` | Group A pedigree companion |

# Troubleshooting {#troubleshooting}

| Problem | Solution |
|------------------------------------|------------------------------------|
| **`package is in use and will not be installed`** | **Session → Restart R**, then reinstall from the clone or GitHub. |
| **Unable to locate `global.R` / `ui.R` / `server.R`** | Reinstall after restart so `inst/shinyapp/` is installed; or `shiny::runApp("E:/ACGM/ClinicalVariantR")`. |
| **Browser does not open** | Copy the `http://127.0.0.1:xxxx` URL from the R console. |
| **VCF not ready / missing CSQ or ANN** | Re-annotate with VEP or SnpEff before upload. |
| **Group A blocked** | Provide clinical logs **and** pedigree CSV in addition to the VCF. |
| **Group C empty results** | Gene symbols must intersect annotated gene names in the VCF. |
| **Locked DLL files (Windows)** | Restart R (`Ctrl+Shift+F10`) before installing or launching. |
| **Large VCF slow on Windows** | Prefer Ubuntu/WSL with bcftools enabled in the analysis options. |
| **Missing Imports (shiny, DT, …)** | Reinstall ClinicalVariantR with `dependencies = TRUE`. |
| **Corrupt lazy-load database** | Restart R, then reinstall with no-staged-install options. |

Interactive launch (copy into an R console; not evaluated here):

```r
library(ClinicalVariantR)
shiny::runApp(ClinicalVariantR(), launch.browser = TRUE)
```

# Session Information

```{r session-info}
sessionInfo()
```
